### 2017-11-29 submission/manuscript.Rmd
grammar updates and flow corrections

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. Next, we assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma. Third, using Random Forest models, we analyzed whether the full community or only the CRC associated genera resulted in better model classification based on the area under the curve (AUC). Since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for.

Our analysis found a continuous decrease in Shannon diversity as disease became more severe which correlated with a significantly increased RR for carcinoma. Using only CRC-associated genera, the RR for carcinoma was higher relative to what was observed for the Shannon diversity RR. Conversely, we demonstrate that the AUC of the classification models increased when the full community was incorporated as opposed to only CRC-associated genera. Although we analyzed data sets which sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.

@@

Within stool, common resident microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6]. Changes in *Bacteroides*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma models [Figure 6]. This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community. These observations would also suggest that initial changes in the resident community, specifically to *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma. Based on these observations, the early changes in the community may give rise to initial polyp formation via interactions with the host. These changes then create new niches in which any one of the CRC-associated genera could gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma via a driver-passenger type mechanism.

The tissue studies did not provide a clearer understanding of how the microbiota may drive the progression of disease severity. For the OTU models both the unmatched and matched [Figure S7E & F] tissue samples had some concordance with the stool data, with resident bacteria being the most prevalent in the top 10 across studies. Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F]. The majority of the results seem to be study specific with many top 10 taxa being present only in a single study. One other potentially worrying sign is the presence of *Propionibacterium* within the top 10 for the genera and OTU models and could be a marker of contamination. The low biomass of these samples coupled with potential contamination might be a possible reason why the tissue results seem to be more sporadic then the stool results.

>>>>

Targeting the identification of biomarkers within stool seems logical since the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not risk free. Stool offers an easy and cost-effective way in which to stratify CRC risk within the population. Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to what is actually happening on the adenoma or carcinoma. Some studies have begun to try and assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Since the stool possibly represents an agregate community of the overall gastrointestinal (GI) tract, sampling the adenoma or carinoma directly could more accurately identify important microbes which could be drastically different then what is identified in stool. Sampling the tissue directly may provide clearer answers but it is not without problems. Due to bowel prep the communities left for sampling may not be reflective of the resident microbiota, but rather a collection of what is able to keep adhered to the mucosa. Additionally, these samples would contain much more host DNA then microbial DNA, potentially limiting the types of analysis that can be done. It has also been well published that low biomass samples can be very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014]. Due to these many differences, one question our meta-analysis aims to answer is whether it is possible to find consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.

This intense focus on identifying biomarkers has identified mostly mouth-associated microbes as potential CRC-associated microbes in the GI tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014]. The main bacteria of interest arising from this set of microbes has been those within the *Fusobacterium* genus. Yet, the question remains as to whether or not this is indeed the most important genera to be focusing on, since many microbiota-based studies typically have identified a collection of oral microbes rather than single species from a single genera [@baxter_microbiota-based_2016; @zeller_potential_2014]. Based on this discrepency, the second question we want to answer with this meta-analysis is if there is one dominant CRC-associated genera that can be identified across studies.

The heavy concentration on the identification of biomarkers has mostly dominated the study of CRC and has had an unattended consequence of reducing the focus on changes that occur within the underlying resident community.  This has been borne out by the majority of previous studies, within stool, tissue, and the only meta-analysis investigating this area, that have focused predominately on biomarker identification. Since it has not been investigated in detail the final question that this meta-analysis aims to answer is whether there are consistent detectable community differences as disease severity increases.

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. We then assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma. Next, using Random Forest models, we analyzed whether the full community or only the CRC-associated genera resulted in better model classification based on the area under the curve (AUC). We observed that the community changes as disease severity worsens and that this community is important for disease classification. However, since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for. Although we analyzed data sets which sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.

@@

Within stool, common resident microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6]. Changes in *Bacteroides*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma models [Figure 6]. This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community. These observations would also suggest that initial changes in the resident community, specifically to *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma. Based on these observations, it is possible to hypothesize that the early changes in the community may give rise to initial polyp formation via interactions with the host and not necessarily via interactions with CRC-assciated genera. These changes then create new niches in which any one of the CRC-associated genera could gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma via a driver-passenger type mechanism.

The tissue studies did not provide a clearer understanding of how the microbiota may drive the progression of disease severity. For the OTU models both the unmatched and matched [Figure S7E & F] tissue samples had some concordance with the stool data, with resident bacteria being the most prevalent in the top 10 important variables across studies. Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F]. The majority of the results seem to be study specific with many top 10 taxa being present only in a single study. One other potentially worrying sign is the presence of *Propionibacterium* within the top 10 for the genera and OTU models and could be a marker of contamination. The low biomass of these samples coupled with potential contamination might be a possible reason why the tissue results seem to be more sporadic then the stool results.

### 2017-11-29 submission/manuscript.tex
grammar updates and flow corrections

The focus on biomarkers within stool makes sense since the current gold
standard for diagnosis is a colonoscopy which can be time-consuming and
is not without risks. Stool offers an easy and cost-effective way in
which to stratify risk within the population. Although stool represents
an easy and less invasive way to assess risk, it is not clear how
reflective this sample is to what is actually happening on the adenoma
or carcinoma. Some studies have begun to try and assess this in health
and disease but are limited by their sample size {[}12,13{]}. Since the
stool possibly represents an agregate community of the overall
gastrointestinal (GI) tract, sampling the adenoma or carinoma directly
might do a better job at identifying important microbes and could be
drastically different then what is identified in stool. Sampling the
tissue directly may provide clearer answers but it is not without
problems. Due to bowel prep the communities left for sampling may not be
the truly resident microbes but rather a collection of what is able to
keep adhered to the mucosa. Further, these samples would contain much
more host DNA then microbial DNA. Low biomass samples are very difficult
to work with and results can end up being study dependent due to the
randomness of contamination {[}14{]}. Due to these variables, one
question our meta-analysis aims to answer is whether it is possible to
find consistent patterns that emerge across studies regardless of
whether they used stool or tissue samples.

This intense focus on identifying biomarkers has identified mostly
mouth-associated microbes as potential CRC-associated microbes in the GI
tract {[}15--17{]}. The main bacteria of interest arising from this set
of microbes has been those within the \emph{Fusobacterium} genus. Yet,
the question remains as to whether or not this is indeed the most
important genera to be focusing on since many studies microbiota based
studies identify a collection of oral microbes rather than single genera
{[}16,17{]}. Based on this discrepency, the second question we want to
answer with this meta-analysis is if there really is one dominant
CRC-associated genera that can be identified across studies.

The heavy focus on the identification of biomarkers has mostly dominated
the study of CRC and has had an unattended consequence of reducing the
focus on changes that occur within the underlying resident community.
This has been borne out by the majority of previous studies, both within
stool and tissue, that have focused predominately on biomarkers as well
as the only meta-analysis investigating this area {[}11{]}. Since it has
not been investigated in detail the final question that this
meta-analysis aims to answer is whether there are consistent detectable
community differences as disease severity increases.

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totalling over 2229 total samples across 14 studies
{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
of the previous meta-analysis to investigate whether the bacterial
community is an important risk factor for both adenoma and carcinoma. To
accomplish this we first assessed whether the diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. Next, we
assessed how common CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
affect the relative risk of adenoma or carcinoma. Third, using Random
Forest models, we analyzed whether the full community or only the CRC
associated genera resulted in better model classification based on the
area under the curve (AUC). We observed that the community changes as
disease severity worsens and that this community is important for
disease classification. However, since the changes in community were
subtle for adenoma we also examined what effect and sample size the
studies that were used were adequately powered for. Although we analyzed
data sets which sampled large numbers of individuals, our results
indicate the individual studies were underpowered for detecting effect
size differences of 10\% or below between the case and control groups.

@@

Within stool, common resident microbes were most consistently present in
the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcus}, and \emph{Roseburia} were
consistently found to be disciminative across the different studies for
both adenoma and carcinoma models {[}Figure 6{]}. This data would
suggest that whether the non-resident bacterium is \emph{Fusobacteria}
or \emph{Peptostreptococcus} is not as important as how these bacteria
interact with the changing resident community. These observations would
also suggest that initial changes in the resident community,
specifically to \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia}, carry on from adenoma to carcinoma. Based on these
observations, the early changes in the community may give rise to
initial polyp formation via interactions with the host. These changes
then create new niches in which any one of the CRC-associated genera
could gain a foothold, exacerbating the initial changes in community and
facilitating the transition from adenoma to carcinoma via a
driver-passenger type mechanism.

The tissue studies did not provide a clearer understanding of how the
microbiota may drive the progression of disease severity. For the OTU
models both the unmatched and matched {[}Figure S7E \& F{]} tissue
samples had some concordance with the stool data, with resident bacteria
being the most prevalent in the top 10 across studies. Unlike in stool,
\emph{Fusobacterium} was the only CRC-associated bacteria consistently
present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
majority of the results seem to be study specific with many top 10 taxa
being present only in a single study. One other potentially worrying
sign is the presence of \emph{Propionibacterium} within the top 10 for
the genera and OTU models and could be a marker of contamination. The
low biomass of these samples coupled with potential contamination might
be a possible reason why the tissue results seem to be more sporadic
then the stool results.

>>>>

Targeting the identification of biomarkers within stool seems logical
since the current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Stool offers an easy and
cost-effective way in which to stratify CRC risk within the population.
Although stool represents an easy and less invasive way to assess risk,
it is not clear how reflective this sample is to what is actually
happening on the adenoma or carcinoma. Some studies have begun to try
and assess this in health and disease but are limited by their sample
size {[}12,13{]}. Since the stool possibly represents an agregate
community of the overall gastrointestinal (GI) tract, sampling the
adenoma or carinoma directly could more accurately identify important
microbes which could be drastically different then what is identified in
stool. Sampling the tissue directly may provide clearer answers but it
is not without problems. Due to bowel prep the communities left for
sampling may not be reflective of the resident microbiota, but rather a
collection of what is able to keep adhered to the mucosa. Additionally,
these samples would contain much more host DNA then microbial DNA,
potentially limiting the types of analysis that can be done. It has also
been well published that low biomass samples can be very difficult to
work with and results can end up being study dependent due to the
randomness of contamination {[}14{]}. Due to these many differences, one
question our meta-analysis aims to answer is whether it is possible to
find consistent patterns that emerge across studies regardless of
whether they used stool or tissue samples.

This intense focus on identifying biomarkers has identified mostly
mouth-associated microbes as potential CRC-associated microbes in the GI
tract {[}15--17{]}. The main bacteria of interest arising from this set
of microbes has been those within the \emph{Fusobacterium} genus. Yet,
the question remains as to whether or not this is indeed the most
important genera to be focusing on, since many microbiota-based studies
typically have identified a collection of oral microbes rather than
single species from a single genera {[}16,17{]}. Based on this
discrepency, the second question we want to answer with this
meta-analysis is if there is one dominant CRC-associated genera that can
be identified across studies.

The heavy concentration on the identification of biomarkers has mostly
dominated the study of CRC and has had an unattended consequence of
reducing the focus on changes that occur within the underlying resident
community. This has been borne out by the majority of previous studies,
within stool, tissue, and the only meta-analysis investigating this
area, that have focused predominately on biomarker identification. Since
it has not been investigated in detail the final question that this
meta-analysis aims to answer is whether there are consistent detectable
community differences as disease severity increases.

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totalling over 2229 total samples across 14 studies
{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
of the previous meta-analysis to investigate whether the bacterial
community is an important risk factor for both adenoma and carcinoma. To
accomplish this we first assessed whether the diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. We then
assessed how common CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
affect the relative risk of adenoma or carcinoma. Next, using Random
Forest models, we analyzed whether the full community or only the
CRC-associated genera resulted in better model classification based on
the area under the curve (AUC). We observed that the community changes
as disease severity worsens and that this community is important for
disease classification. However, since the changes in community were
subtle for adenoma we also examined what effect and sample size the
studies that were used were adequately powered for. Although we analyzed
data sets which sampled large numbers of individuals, our results
indicate the individual studies were underpowered for detecting effect
size differences of 10\% or below between the case and control groups.

@@

Within stool, common resident microbes were most consistently present in
the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcus}, and \emph{Roseburia} were
consistently found to be disciminative across the different studies for
both adenoma and carcinoma models {[}Figure 6{]}. This data would
suggest that whether the non-resident bacterium is \emph{Fusobacteria}
or \emph{Peptostreptococcus} is not as important as how these bacteria
interact with the changing resident community. These observations would
also suggest that initial changes in the resident community,
specifically to \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia}, carry on from adenoma to carcinoma. Based on these
observations, it is possible to hypothesize that the early changes in
the community may give rise to initial polyp formation via interactions
with the host and not necessarily via interactions with CRC-assciated
genera. These changes then create new niches in which any one of the
CRC-associated genera could gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma via a driver-passenger type mechanism.

The tissue studies did not provide a clearer understanding of how the
microbiota may drive the progression of disease severity. For the OTU
models both the unmatched and matched {[}Figure S7E \& F{]} tissue
samples had some concordance with the stool data, with resident bacteria
being the most prevalent in the top 10 important variables across
studies. Unlike in stool, \emph{Fusobacterium} was the only
CRC-associated bacteria consistently present in the top 10 of the CRC
models {[}Figure S7B-C \& E-F{]}. The majority of the results seem to be
study specific with many top 10 taxa being present only in a single
study. One other potentially worrying sign is the presence of
\emph{Propionibacterium} within the top 10 for the genera and OTU models
and could be a marker of contamination. The low biomass of these samples
coupled with potential contamination might be a possible reason why the
tissue results seem to be more sporadic then the stool results.
